---
title:  "Review the basics"
type:   assignment
---

```{r echo=FALSE, include=FALSE}
library(courseR)
library(checkr)
```

# R 101 review

Let's start simple and review the use of variables and doing basic math operations.

Save the number `10` in a variable named `a` and then print its contents:

```{r solution}
a <- 10
a
```

```{r checkr}
find_call("a <- 10", "You'll need to use the assignment operator ('<-').")
check_final(match_number(10), "Uh, it should be 10.")
```

Store `2` in the variable `b` and print it:

```{r solution}
b <- 2
b
```

```{r checkr}
find_call("b <- 2", "You'll need to use the assignment operator ('<-').")
check_final(match_number(2), "Uh, it should be 2.")
```

Add `a` to `b`:

```{r solution}
a + b
```

```{r checkr}
find_call("a + b", "You'll need to use the addition operator ('+').")
```

Subtract `b` from `a`:

```{r solution}
a - b
```

```{r checkr}
find_call("a - b", "You'll need to use the subtraction operator ('-').")
```

Multiply `a` and `b`:

```{r solution}
a * b
```

```{r checkr}
find_call("a * b", "You'll need to use the multiplication operator ('*').")
```

Divide `a` by `b` (surprise, surprise):

```{r solution}
a / b
```

```{r checkr}
find_call("a / b", "You'll need to use the division operator ('/').")
```

Raise `a` to the power of `b`:

```{r solution}
a^b
```

```{r checkr}
find_call("a^b", "You'll need to use the power operator ('^').")
check_final(match_number(a^b), "Uh, it should be 10.")
```

# Strings

Store the strings "Hello" and "world" in a character vector assigned to the variable `welcome` and then evaluate `welcome`:

```{r solution}
welcome <- c("Hello", "world")
welcome
```

Find the `length` of your `welcome` variable:

```{r solution}
length(welcome)
```

```{r checkr}
find_call("length(welcome)", "You'll want to use the `length()` function.")
```

# Practice Indexing

Create a vector holding the numbers `1` through `42` and save it in a variable called `x`. Then print it.

```{r solution}
x <- 1:42
x
```

```{r checkr}
find_call("1:42", "The easiest way to do this is to use the colon operator (':').")
```

Extract the numbers in this vector at positions 10 and 20 (in the same operation):

```{r solution}
x[c(10, 20)]
```

Extract numbers 10 through 20:

```{r solution}
x[10:20]
```

```{r checkr}
find_call("10:20", "The easiest way to do this is to use the colon operator (':').")
```

Create a vector containing the remainder when each number in `x` is divided by 2:

```{r solution}
x %% 2
```

```{r checkr}
find_call("x %% whatever", "The easiest way to do this is to use the modulo operator ('%%').")
```

Create a logical vector which is `TRUE` for every numbers and `FALSE` for odd numbers in `x`:

```{r solution}
x %% 2 == 0
```

```{r checkr}
find_call("x %% 2", "The easiest way to do this is to use the modulo operator ('%%').")
find_call("whatever == whatever", "You'll want to use the comparison operator to compare values ('==').")
```

Now, put it all together to extract the even numbers from `x`:

```{r solution}
x[x %% 2 == 0]
```

Replace all of the odd numbers in the original `x` vector with the missing value flags (`NA`), then print `x`.

```{r solution}
x[x %% 2 == 1] <- NA
x
```

