---
title:  "Visualizing Data"
type:   content
---

# Introduction

Today you're going to learn how to make this:

```{r warning=FALSE}
library(ggplot2)
ggplot(diamonds, aes(depth, fill = cut)) + geom_density(alpha = 0.2) + xlim(55, 70)
```

It's only going to take one line of code, but there are a few things we need to cover to understand it!

# Tables: data.frame's

##

One of the data sets that `ggplot2` comes with is `diamonds`.  Get a preview:

```{r eval=FALSE}
diamonds
```

```{r echo=FALSE}
head(diamonds)
```

The `diamonds` variable holds a new type of data we haven't seen before: this is a `data.frame` (well, technically it's a [tibble](https://github.com/tidyverse/tibble) but we're going to ignore that distinction for now).  This is the data structure in R that you'll nearly always use to represent a real data set: it's two dimensional, rows represent observations, and each column represents a variable.

##

You can see how many observations a data set has with `nrow`:

```{r}
nrow(diamonds)
```

That's a few!

##

Each column on a `data.frame` is just a vector of values.  To get the vector of data in a column you can use the short hand `$` syntax:

```{r eval=FALSE}
diamonds$cut
```

```{r echo=FALSE}
head(diamonds$cut)
```

##

You can also use the longer form `[[..]]` syntax:

```{r eval=FALSE}
diamonds[['cut']]
diamonds[['cut']] == diamonds$cut
```

##

Note the use of quotes with `[[..]]` but not with `$`.  See what happens if you use single `[...]` on a `data.frame` column:

```{r eval=FALSE}
diamonds['cut']
```

```{r echo=FALSE}
head(diamonds['cut'])
```

What did you get?  What's the difference between double and single brackets?

##

RStudio also has a very nice interface for inspecting a `data.frame` with the `View` function (note caps):

```{r eval=FALSE}
View(diamonds)
```

# Plotting

We can make basic plots with the `plot` function.  Try passing `plot` two quantitative variables:

```{r}
plot(diamonds$carat, diamonds$price)
```

What did you get?

##

Or a quantitative and categorical variable:

```{r}
plot(diamonds$cut, diamonds$price)
```

What did you get there?

The `plot` function is a little bit magical.  It will try very, very hard to draw something no matter what you pass it.  But it doesn't always choose well (if you want to crash your session run `plot(diamonds)`) and you don't even want to know what you have to do to customize the output and make the plots pretty (see `?par`).  It's an awesome function to try out when you're first playing around with a new package or type of data, but you'll quickly be ready to move onto something higher-level.  That's nearly always going to be `ggplot2`.

The `ggplot2` package implements a [semantic visualization](http://vita.had.co.nz/papers/layered-grammar.pdf) framework.  The idea here is to be systematic about separating and independently describing the fundamental components of a visualization.  If you've ever been frustrated with the inscrutable choices that Excel makes when you try to make a plot, you'll appreciate why this is important.

##

The first component of any plot is **the data**.  You define the table by passing it as the first argument to the `ggplot2` function:

```{r}
ggplot(diamonds)
```

And you get a beautiful empty box.  Exciting!  We'll get there.

The second component of any plot is **what** you are going to visualize: which variables are you using and what relationship are you interested in?  You do this by passing an "aesthetic mapping" as the second value to `ggplot` with the `aes` function.

##

Visualize the `price` variable:

```{r}
ggplot(diamonds, aes(price))
```

##

Visualize the `cut` variable:

```{r}
ggplot(diamonds, aes(cut))
```

##

Visualize `price` as a function of `carat` (independent, dependent):

```{r}
ggplot(diamonds, aes(carat, price))
```

Why does this relationship make more sense than the inverse?

Easy enough, but we're still just getting a grey boxes.  To actually draw something, we need to define the last component of any visualization: **how** the variables should be projected into an image.  In `ggplot` that's the job of the "geom_" functions and well add visualizations to a plot using the `+` operator:

##

A scatter plot with **point**s:

```{r}
ggplot(diamonds, aes(carat, price)) + geom_point()
```

##

A histogram:

```{r}
ggplot(diamonds, aes(price)) + geom_histogram()
```

##

A density distribution:

```{r}
ggplot(diamonds, aes(price)) + geom_density()
```

How do these two methods compare for visualizing the dispersion of values in a quantitative variable?

<br>

# Grouping variables

Believe it or not we're really close to creating the plot at the start of the document.  In it, we're visualizing the density distribution of the `depth` variable:

```{r}
ggplot(diamonds, aes(depth)) + geom_density()
```

But in the plot above values for `depth` are grouped by the categorical variable `cut`.  Specifically, the density traces are "filled" with difference colors based on `cut`.

##

To use fill color to group values, we add that to the aesthetic mapping, because it's about **what** we're plotting:

```{r}
ggplot(diamonds, aes(depth, fill = cut)) + geom_density()
```

##

Getting close, but there's a lot of overlap between the traces, so it would be nice to make them a little transparent.  Do that with the `alpha` argument; it's about the **how** the plot is drawn:

```{r}
ggplot(diamonds, aes(depth, fill = cut)) + geom_density(alpha = 0.2)
```

##

Finally, to focus in on the main body of the data (binning out very large and very small values), we can set the x-axis limits with `xlim`:

```{r}
ggplot(diamonds, aes(depth, fill = cut)) + geom_density(alpha = 0.2) + xlim(55, 70)
```

And there it is!
