---
title:  "Review the basics"
---

# R 101 review

Let's start simple and review the use of variables and doing basic math operations.

Save the number `10` in a variable named `a` and then print its contents:

```{r}

```

Store `2` in the variable `b` and print it:

```{r}

```

Add `a` to `b`:

```{r}

```

Subtract `b` from `a`:

```{r}

```

Multiply `a` and `b`:

```{r}

```

Divide `a` by `b` (surprise, surprise):

```{r}

```

Raise `a` to the power of `b`:

```{r}

```

# Strings

Store the strings "Hello" and "world" in a character vector assigned to the variable `welcome` and then evaluate `welcome`:

```{r}

```

Find the `length` of your `welcome` variable:

```{r}

```

# Practice Indexing

Create a vector holding the numbers `1` through `42` and save it in a variable called `x`. Then print it.

```{r}

```

Extract the numbers in this vector at positions 10 and 20 (in the same operation):

```{r}

```

Extract numbers 10 through 20:

```{r}

```

Create a vector containing the remainder when each number in `x` is divided by 2:

```{r}

```

Create a logical vector which is `TRUE` for even numbers and `FALSE` for odd numbers in `x`:

```{r}

```

Now, put it all together to extract the even numbers from `x`:

```{r}

```

Replace all of the odd numbers in the original `x` vector with the missing value flags (`NA`), then print `x`.

```{r}

```

