---
title:  "Review the basics"
type:   assignment
---

# Working on assignments

As you are working you can Knit your assignment files like any other R Markdown file.  When you're ready to check your answers against an answer key, run:

```{r eval=FALSE}
bio185::checkAssignment("01-the-basics")
```

The course app will launch in the RStudio viewer pane.  If you change your source R Markdown file while the app is open, it will automatically try to re-Knit it and update if it succeeds.

When you're done, close the course app by hitting the "Stop" button if it's still running, and then call:

```{r eval=FALSE}
bio185::submitAssignment("01-the-basics")
```

You can re-submit assignments as many times as you like.

**Note**:  The course app is a brand new piece of software and there are likely to be bugs!  Let me know when you have problems.

<br>

# Review the basic basics

Practice doing some math!  Create a few vectors (of different lengths) and show that you remember how to:

* add/subtract
* multiply/divide
* raise to an exponent

The answer key is going to use these vectors:

```{r}
a <- c(1, 2, 3)
b <- c(4, 5, 6)
```

```{r solution}
a + b
a - b
a * b
a / b
a ^ b
```

Store the strings "Hello" and "world" in a character vector assigned to the variable `welcome` and then evaluate `welcome`:

```{r solution}
welcome <- c("Hello", "world")
welcome
```

Find the `length` of your `welcome` variable:

```{r solution}
length(welcome)
```

<br>

# Indexing vectors

For all the following questions the answer key is going to use this vector:

```{r}
c <- 1:42
```

Extract the numbers in this vector at positions 10 and 20 (in the same operation):

```{r solution}
c[c(10, 20)]
```

Extract numbers 10 through 20:

```{r solution}
c[10:20]
```

Create a numeric vector testing the remainder when each number in this vector is divided by 2 (*hint*: remember the `%%` operator):

```{r solution}
c %% 2
```

Create a logical vector which is `TRUE` for every even number and `FALSE` for odds:

```{r solution}
c %% 2 == 0
```

Extract the even numbers from our vector above:

```{r solution}
c[c %% 2 == 0]
```

Replace all of the odd numbers in the original vector with the missing value `NA`:

```{r solution}
c[c %% 2 == 1] <- NA
c
```

<br>

# Basic plotting with ggplot2

Load the `ggplot2` package:

```{r solution}
library(ggplot2)
```

We'll use the `diamonds` data set for the rest of the questions in this section; remind yourself what variables are available by printing a preview of the table:

```{r solution}
diamonds
```

Make a histogram showing the distribution of the `price` variable:

```{r solution}
ggplot(diamonds, aes(price)) + geom_histogram()
```

Make a density distribution showing the same:

```{r solution}
ggplot(diamonds, aes(price)) + geom_density()
```

Does diamond color affect prices?  Do the same as above, but group prices by the color of the diamond; optionally make it pretty using a little transparency:

```{r solution}
ggplot(diamonds, aes(price, fill = color)) + geom_density(alpha = 0.2)
```

Let's end with a bit more of a challenge; you'll probably need to consult the `ggplot2` documentation with `help` to do this one.  Plot the price of diamonds as a function of carats (why does this relationship make more sense than the inverse?) with a scatter plot.  Color the points based on the cut of the diamond.  Use transparency to make it easier to see overlapping points.

```{r solution}
ggplot(diamonds, aes(carat, price, color = cut)) + geom_point(alpha = 0.3)
```

