---
title:  "Making plots"
type:   assignment
data:   recent-grads.csv
---

# Data

For this assignment we'll continue working with the [FiveThirtyEight college majors](https://github.com/fivethirtyeight/data/tree/master/college-majors) data set we worked with in class.  In class you loaded the data into a variable named `recent_grads` using the RStudio import wizard.  For this assignment, I've given you a local copy in your `data/` folder.  Remember that you can review what data are held in [each column here](https://github.com/fivethirtyeight/data/tree/master/college-majors).

Here's how to load the table from a file (just leave this block in place):

```{r}
recent_grads <- readr::read_csv("data/recent-grads.csv")
```

Like we did in class, convert the `Major_category` column from a `character` fector into a `factor` vector.

```{r solution}
recent_grads$Major_category <- factor(recent_grads$Major_category)
```

Use `summary` to make sure the data look good:

```{r solution}
summary(recent_grads)
```

<br>

# Exploring the data set

For each of the following questions, build a plot that is a useful way to visualize an answer.  It's ok if your plots aren't identical to mine; there are multiple "correct" solutions for each.

What is the overall distribution of median incomes?  What major is the extreme outlier on the high-end (click to sort columns in the RStudio table viewer)?

```{r solution}
ggplot(recent_grads, aes(Median)) + geom_density()
```


Is there a correlation between the percentage of women majors and median post-graduate income?  Are some major categories more skewed than others?

```{r solution}
ggplot(recent_grads, aes(ShareWomen, Median, color = Major_category)) + geom_point()
```

Is there a correlation between median income and unemployment rate?  Is there any skewing with respect to the percentage of women in the field?

```{r solution, warning=FALSE}
ggplot(recent_grads, aes(Median, Unemployment_rate, color = ShareWomen)) + geom_point()
```

What is the distribution of median incomes across the major categories?  *Hint*:  see `geom_boxplot`.  *Pro-tip*: to change the angle of your x-axis labels, you add something like this to your plot `theme(axis.text.x = element_text(angle = 60, hjust = 1))`.

```{r solution}
ggplot(recent_grads, aes(Major_category, Median)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

